---
title: "Enhancing Your ggplot"
subtitle: "Statistics Homework Extension"
format:
  html:
    code-fold: true
    theme: cosmo
  pdf:
    documentclass: article
---

## Overview

In this assignment, you’ll build on your earlier bar chart by learning how to **enhance your ggplot visualizations** with color, labels, themes, and other styling options.

You’ll start with a simple plot and gradually add improvements to make it more clear and visually appealing.

---

## Starter Code

Begin with the basic bar chart you created in class.

```{r}
library(ggplot2)
library(tidyverse)

data <- data.frame("Categories" = c("A", "B", "C", "D"), 
                   "Values" = c(1, 3, 6, 2))

ggplot(data, aes(x = Categories, y = Values)) + 
    geom_col()
```

### Add color to bars 
You can control the fill color and the outline color

```{r}
ggplot(data, aes(x = Categories, y = Values)) +
    geom_col(fill = "skyblue", color = "black")
```

* `fill` changes the color inside the bars 
* `color` changes the border color of each bar 

### Add axis labels and a title 

Use `labs()` to label axes and add a title 

```{r}
ggplot(data, aes(x = Categories, y = Values)) +
    geom_col(fill = "lightblue", color = "black") +
    labs(title = "My Sample Bar Chart",
         x = "Category Type",
         y = "Measured Value")
```

* `title`, `x`, `y` change the labels of the corresponding elements 

### Change the theme 

You can use built-in ggplot themes to change background and grid styles 

```{r}
ggplot(data, aes(x = Categories, y = Values)) +
    geom_col(fill = "lightblue", color = "black") +
    labs(title = "My Sample Bar Chart",
         x = "Category Type",
         y = "Measured Value") +
    theme_minimal()
```

* `theme_classic()` - removes gridlines 
* `theme_bw()` - black-and-white 
* `theme_dark()` - dark background 

### Adjust font and text size 

You can further customize `theme()` to modify how texts appears 

```{r}
ggplot(data, aes(x = Categories, y = Values)) +
    geom_col(fill = "lightblue", color = "black") +
    labs(title = "My Sample Bar Chart",
         x = "Category Type",
         y = "Measured Value") +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)
    )
```

* `plot.title` controls size, style, and alignment 
* `hjust = 0.5` centers the title 
* `axis.title.x/y` adjust axis label text

### Add labels to each bar 

Use `geom_text()` to label each bar with its value 

```{r}
ggplot(data, aes(x = Categories, y = Values)) +
    geom_col(fill = "lightblue", color = "black") +
    geom_text(aes(label = Values), vjust = -0.3) +
    labs(title = "My Sample Bar Chart",
         x = "Category Type",
         y = "Measured Value") +
    theme_minimal()
```

* `geom_text()` adds numeric labels 
* `vjust` adjusts how high or low the text appears (negative = below bar) 

### Challenge 

Put it all together to create your own best version of a bar chart. Use the following dataset to create a plot. 

```{r}
data <- iris %>% filter(Species == "setosa")
```

