---
title: "2.2 Homework"
execute: 
  message: false
  warning: false
---

## Setup 
Before running any future code, run the following chunk. This will check if the required packages are installed, load them if they are, and install them if they are not. 

```{r setup}
if (!require("tidyverse")) {
    install.packages("tidyverse")
    library(tidyverse)}
if (!require("ggplot2")) {
    install.packages("ggplot2")
    library(ggplot2)}
```

## Functions  
To work through the following exercises, copy the code block contents required for the given problem to the code block provided and enter the data the problem gives. 

### Histogram 
A histogram shows how numerical data are distributed across intervals (bins). Use `geom_histogram()` in `ggplot()` to create one.

```{r histogram example}
# Example using manufactured data

data <- data.frame("Heights" = c(58, 60, 61, 62, 63, 64, 64, 65, 65, 66, 67, 67, 68, 68, 69, 69, 70, 71, 71, 72))

ggplot(data, aes(x = Heights)) +
    geom_histogram(binwidth = 2, color = "black", fill = "skyblue") +
    labs(title = "Histogram of Student Heights",
         x = "Height (inches)",
         y = "Frequency")
```

### Frequency Polygon 
A frequency polygon is similar to a histogram but uses lines to connect the midpoints of intervals.
You can use `geom_freqpoly()` with the same binwidth as a histogram.

```{r Freqpoly}
data <- data.frame("Test_Scores" = c(55, 57, 58, 60, 61, 62, 63, 64, 65, 67, 68, 70, 72, 73, 75, 77, 78, 80, 82, 85))

ggplot(data, aes(x = Test_Scores)) +
    geom_freqpoly(binwidth = 5, color = "darkred", size = 1) +
    labs(title = "Frequency Polygon of Test Scores",
         x = "Score",
         y = "Frequency")
```

### Time Series 
A time series graph shows how a quantity changes over time. Use a line graph, but with time (or an ordered variable) on the x-axis. 

```{r}
# Example using manufactured data

data <- data.frame(
    Day = 1:10,
    Visitors = c(120, 135, 140, 160, 180, 170, 165, 190, 195, 210)
)

ggplot(data, aes(x = Day, y = Visitors)) +
    geom_line(color = "blue") +
    geom_point() +
    labs(title = "Daily Visitors to Museum Over 10 Days",
         x = "Day",
         y = "Number of Visitors")

```

## Problems 
### Problem 1 - Histogram 
The following data show the number of minutes 30 students spent on social media yesterday:
32, 34, 36, 38, 40, 41, 43, 44, 44, 45, 46, 48, 49, 49, 50, 51, 52, 53, 55, 55, 56, 58, 59, 61, 63, 64, 65, 67, 68, 69

Create a histogram of the data using a binwidth of 5.

```{r P1}

```

### Problem 2 - Histogram 
The data show the daily high temperature (°F) in a city over 30 days:
61, 62, 62, 63, 64, 65, 65, 66, 66, 67, 68, 68, 68, 69, 70, 70, 71, 72, 72, 73, 74, 74, 75, 75, 76, 76, 77, 78, 79, 80

Create a histogram with a binwidth of 2. 

```{r P2}

```

### Problem 3 - Frequency Polygon 
Create a frequency polygon of the daily high temperatures (°F) recorded over 20 days:
62, 63, 63, 65, 67, 68, 68, 69, 70, 71, 71, 72, 73, 74, 75, 75, 76, 77, 78, 80

Create a frequency polygon of this data.

```{r P3}

```

### Problem 4 - Frequency Polygon 
Create a frequency polygon of reaction times (in milliseconds) from a group of test subjects:
180, 190, 195, 200, 205, 210, 215, 220, 225, 230, 235, 240, 245, 250, 255, 260, 265, 270, 275, 280

Create a frequency polygon with clearly labeled axes.

```{r P4}

```

### Problem 5 — Time Series

A company recorded its monthly online sales for one year:
Month: Jan–Dec
Sales (in $1000): 20, 25, 28, 30, 35, 40, 38, 37, 42, 45, 50, 48

Create a time series graph showing sales over time. 

```{r}
data <- data.frame(
    Month = factor(month.abb, levels = month.abb),
    Sales = c(20,25,28,30,35,40,38,37,42,45,50,48)
)

ggplot(data, aes(x = Month, y = Sales, group = 1)) +
    geom_line(color = "steelblue") +
    geom_point() +
    labs(title = "Monthly Online Sales",
         x = "Month",
         y = "Sales (in $1000)")

```

### Problem 6 - Time Series 
The following data show the average number of daily visitors to a local museum over 10 days:
Days: 1–10
Visitors: 120, 135, 140, 160, 180, 170, 165, 190, 195, 210

Create a time series plot of this data.

```{r P6}

```

### Challenge Problem (Optional)

Create your own small dataset showing how something changes over time — for example, the number of hours you studied each day for a week.
Use a time series plot to display your data. Be sure to label your axes and include a descriptive title. 

```{r Challenge}

```

