---
title: "2.2 - Histograms, Frequency Polygons, and Time Series Graphs"
execute: 
    echo: false
---


```{r setup}
#| echo: false
#| warning: false
#| message: false

source("book_functions.R")
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gt)

```

## Histograms 

* Much of the work in this book is displayed with histograms. 
* Advantage: can readily display large data sets 
* Consists of contiguous boxes, a horizontal axis (more or less a number line), and a vertical axis labeled with **frequency** or **relative frequency**. 
    * Relative frequency is equal to frequency of observed value divided by total number of data values 
    * If $f=$ frequency, $n=$ total number of data values, and $RF=$ relative frequency, then $RF=\frac{f}{n}$ 
* Shows shape of the data - normal, skewed, and spread 

### Constructing a histogram 

* To construct a histogram, decided how many **bars** or **intervals** you want 
* The width of a bar is called bin size, found dividing the range of the data by desired number of bins 
* Number of observations in each bin = frequency 

We won't need to construct a histogram by hand in this class, instead we can use RStudio to help us. 

### Example 1 

The following data are the heights (in inches, to nearest half inch) of 100 male semiprofessional soccer players. The heights are **continuous** since height is measured. 

The following histogram displays the height on the $x$-axis and the relative frequency on the $y$-axis. 

```{r}
x <- c(60, 60.5, 61, 61, 61.5, 63.5, 63.5, 63.5, 64, 64, 64, 64, 64, 64, 64, 64.5, 64.5, 64.5, 64.5, 64.5, 64.5, 64.5, 64.5, 66, 66, 66, 66, 66, 66, 66, 66, 66, 66, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 66.5, 67, 67, 67, 67, 67, 67, 67, 67, 67, 67, 67, 67, 67.5, 67.5, 67.5, 67.5, 67.5, 67.5, 67.5, 68, 68, 69, 69, 69, 69, 69, 69, 69, 69, 69, 69, 69.5, 69.5, 69.5, 69.5, 69.5, 70, 70, 70, 70, 70, 70, 70.5, 70.5, 70.5, 71, 71, 71, 72, 72, 72, 72.5, 72.5, 73, 73.5, 74)

bin_breaks <-  seq(59.95, 75.95, by=2)

data.frame(x=x) %>% 
    ggplot(aes(x=x)) + 
    geom_histogram(aes(y = after_stat(count)/sum(after_stat(count))),breaks = bin_breaks, color = "black", fill = "lightgray") + 
    theme_notes() + 
    scale_x_continuous(breaks=bin_breaks)
```

## Frequency Polygons 

* Analogous to line graphs 
* Good for overlaying data
* Data is divided into a number of intervals (think bins from histograms), giving $x$ values 
* Find frequency of data within each interval, giving $y$ values 
* Plot points for each interval, draw line segments to connect them. 

### Example 2 

The following frequency polygon was constructed by looking at the distribution of Calculus final test scores. 

```{r}
ggplot(iris, aes(x = Petal.Length)) + 
    geom_freqpoly(bins = 10) + 
    geom_point(stat = "bin", aes(y = after_stat(count)), bins = 10) +
    theme_notes(show_axis_titles = TRUE) + 
    labs(x = "Petal Length", 
         y = "Frequency", 
         title = "Length of Iris Petals")
```

Can separate out different data categories and lay on top: 

```{r}
ggplot(iris, aes(x = Petal.Length, color = Species)) + 
    geom_freqpoly(bins = 10) + 
    geom_point(stat = "bin", aes(y = after_stat(count)), bins = 10) +
    theme_notes(show_axis_titles = TRUE) + 
    labs(x = "Petal Length", 
         y = "Frequency", 
         title = "Length of Iris Petals")
```

## Time Series Graph 

* Suppose we want to study temperature of a region for an entire month. 
* One feature we may want to consider is time. Since date is paired with temperature, we can use time as an axis. 
* We use both pieces of the paired data to construct the graph - one axis for time, one axis for data value. Typically points are connected by lines in order of occurrence. 

### Example 3 

The following time series shows the number unemployed (in thousands) on the $y$-axis, and the date on the $x$-axis. 

```{r}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(title = "Unemployment in the U.S. (1967–2015)",
       x = "Year", y = "Number unemployed (thousands)") +
  theme_minimal(base_size = 14)

```

