# 5.1 Continuous Probability Functions

-   Defined as a continuous function $f(x)$ such as a uniform, exponential, or normal function.
-   Functions are defined so that the area between the function and the $x$-axis is equal to a probability.
    -   Since max probability is 1, the area under the curve must be 1.
-   **Remember, for continuous probability functions, PROBABILITY = AREA**

## Examples

Consider the function $f(x) = 1/20$ for $0 \le x \le 20$ where $x$ is a real number.

```{r}
#| echo: false
#| label: fig-plot1
#| fig-cap: "$f(x) = 1/20$"
#| fig-width: 4
#| fig-height: 3
#| message: false
library(ggplot2)
labels <- function(x) {
  #paste(x, "grams")
    MASS::fractions(x)
}
df <- data.frame(x = c(0,20,0), y = c(0, 0,0.05), xend = c(0, 20, 20), yend = c(0.05, 0.05, 0.05))
ggplot() + 
    theme_minimal(base_size = 14) + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black")) + 
    xlim(0, 22) + 
    geom_segment(data = df, aes(x = x, y = y, xend = xend, yend = yend)) + 
    labs(y = "f(x)") +
    geom_area(data = data.frame(x = c(0, 20), y = c(0.05, 0.05)), aes(x = x, y = y),fill = "lightblue", alpha = .5) +
    scale_y_continuous(label = labels, limits = c(0, .1)) + 
    scale_x_continuous(breaks = c(0, 20))
```

The area under the curve for this function is the area of a rectangle: $Area = b \times h = 20 \times \frac{1}{20} = 1$ So the total probability is 1. We can now find the probability of *ranges* of $x$-values as the area of different rectangles under the curve.

**Suppose we want to find the area instead where** $0<x<3$ instead:

```{r}
#| label: fig-plot2
#| echo: false
#| message: false
#| fig-cap: "$f(x) = 1/20 \\text{ for } 0 \\leq x \\leq 3$"
#| fig-height: 3
#| fig-width: 4

ggplot() + 
    theme_minimal(base_size = 14) + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black")) + 
    xlim(0, 22) + 
    geom_segment(data = df, aes(x = x, y = y, xend = xend, yend = yend)) + 
    labs(y = "f(x)") +
    geom_area(data = data.frame(x = c(0, 20), y = c(0.05, 0.05)), aes(x = x, y = y),fill = "lightblue", alpha = .5) + 
    geom_area(data = data.frame(x = c(0, 3), y = c(0.05, 0.05)), aes(x = x, y = y),fill = "blue", alpha = .7) + 
    scale_y_continuous(label = labels, limits = c(0, .1)) + 
    scale_x_continuous(breaks = c(0, 3, 20))
```

As you can see in @fig-plot2, the shaded region is a rectangle with base 3 ($0 - 3$) and height $\frac{1}{20}$ and $Area = 3 \cdot \frac{1}{20} = \frac{3}{20} = .15$

This area corresponds to probability $P(0<x<3) = 0.15$

**Similarly, suppose we want to find the area where** $4<x<15$:

```{r}
#| label: fig-plot3
#| echo: false
#| message: false
#| fig-cap: "$f(x) = 1/20 \\text{ for } 4 \\leq x \\leq 15$"
#| fig-height: 3
#| fig-width: 4

ggplot() + 
    theme_minimal(base_size = 14) + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black")) + 
    xlim(0, 22) + 
    geom_segment(data = df, aes(x = x, y = y, xend = xend, yend = yend)) + 
    labs(y = "f(x)") +
    geom_area(data = data.frame(x = c(0, 20), y = c(0.05, 0.05)), aes(x = x, y = y),fill = "lightblue", alpha = .5) + 
    geom_area(data = data.frame(x = c(4, 15), y = c(0.05, 0.05)), aes(x = x, y = y),fill = "blue", alpha = .7) + 
    scale_y_continuous(label = labels, limits = c(0, .1)) + 
    scale_x_continuous(breaks = c(0, 4, 15, 20))
```

From @fig-plot3, the shaded region is a rectangle with base 11 ($15 - 4$) and height $\frac{1}{20}$ and $Area = 11 \cdot \frac{1}{20} = \frac{11}{20} = .55$.

This area corresponds to probability $P(4<x<15) = 0.55$

### Cumulative Distribution Function (CDF)

$P(X \le x)$ is called the **Cumulative Distribution Function** or CDF. This is the probability that $X$ is less than $x$. CDF gives the *area to the left* and $P(X > x)$ *gives area to the right*. These are related as: $P(X>x) = 1-P(X<x)$

### Question 1

What is the probability that $x$ takes a value between 2.3 and 12.7?

::: {pdf}
\vspace{2cm}
:::

:::: {html}
::: {style="height: 50px;"}
:::
::::

### Question 2

Consider the function $f(x) = 1/6$ for $0 \le x \le 6$. Draw the graph of $f(x)$ and find $P(2.5 \le x \le 5.5)$
