---
title: "Mini Data Visualization Project Instructions"
author: "Your Name"
format:
  html:
    embed-resources: true
    theme: cosmo
  pdf:
    documentclass: article
    geometry: margin=1in
execute:
    echo: true
    warning: false
    message: false
---

## Prelude: 

Remember to put save this file into an R specific folder that you know where it is so you can easily find it! 

## Setup 

Load any packages you might need by putting them in the following block and running it: 

```{r setup}
# Example: 
library(ggplot2)
# Place any other packages you need here: 

```

## Choose your dataset 

Pick one of the following: 

* A built-in R dataset such as
    * `mtcars`
    * `iris` 
    * `penguins` 
    * `ToothGrowth` 
    * `diamonds` 
    * `storms` 
    * or another set found at `library(help = "datasets")` 
* Or a small CSV dataset of your own (check with me that it will work) 

If you'd like to explore the above datasets, run `head(data_set_name)` in the Console below. 

```{r dataset}
# Using a built in dataset: 
data <- penguins

# Or use your own
# Uncomment and modify:
# data <- read.csv("your_data.csv")

# View the first few rows of your chosen data: 

head(data)
```


## Create a graph 
Choose one graph type we learned about in class: 

* Stem-and-leaf: `stem()` 
* Bar graph: `geom_bar()` 
* Histogram: `geom_histogram()` 
* Box plot: `geom_boxplot()` 
* Time series line graph: `geom_line()` 

Your graph should include: 

* A title 
* Labeled axes 
* Readable scales 
* Appropriate bin widths (for histograms) 

```{r graph}
# Place your graph here 
# Check prior homework assignments for layout 

```


## Interpret your graph 
Write a short (3-6) sentence paragraph that includes: 

* What variable(s) you plotted 
* What the overall pattern or shape looks like 
* Any clusters, outliers, or unusual features 
* What the graph tells you about the data 

Interpretation: 

## Knit and submit 

* Save your Quarto document as `lastname_vizproject.qmd` 
* Knit to HTML (or PDF) and submit them to me via email. 